---
title: "20240817 MAIT characterisation MAIT clustering for thesis"
author: "Abbey Figliomeni"
date: "2024-08-17"
output: html_document
---

# Data import and preprocessing
## Extract flow data (class: gatingset) using cytoML
```{r, results="hide"}
source("[cytoML script]")
```

## Libraries
```{r, results='hide', message=FALSE}
library(Spectre)
Spectre::package.check()    # Check that all required packages are installed
Spectre::package.load()  
```

## Convert gatingset object to cytoset
Cytoset object extracted from gatingset, then subsetted using flowJo "MAIT cell" gate.
```{r}
cs <- gs_pop_get_data(gs, "MAIT cells")
cs  # returns a cytoset with only cells contained within MAIT 
```

## Add meta data, convert to data.table, and merge files into list
```{r}
### create list of converted data.tables with phenodata
data.list <- list()  # creates empty list for data.tables
pDatalist <- colnames(pData(cs))  # column names from cs phenodata as a list


### convert and append cytoframes 
for(s in 1:length(sampleNames(cs))){
  a <- exprs(cs[[s]])  # get extract flowframe data as matrix
  b <- as.data.frame(a)  #convert flowframe data to dataframe
  col_change_vec <- grep("Comp", colnames(cs[[s]]))  #creates numerical vector with which colnames contain "comp" (ie not FSC SSC Time)
  y <- keyword(cs[[s]])  # extract keywords from flowframe (column numbers 'n' in flowframe correspond to $Pn, parameter names under $PnS)
  for(n in col_change_vec){
    colnames(b)[[n]] <- y[[paste("$P", n, "S", sep="")]]  # finds matching keyword for columns containing comp parameters
  }
  b$`FSC-A` <- NULL
  b$`FSC-H` <- NULL
  b$`FSC-W` <- NULL
  b$`SSC-A` <- NULL # delete irrelevant parameters (could do this with vector instead of one at a time)
  b$`SSC-H` <- NULL
  b$`SSC-W` <- NULL
  b$Time <- NULL
  b$ID <- sampleNames(cs[s])  #creates ID column corresponding to FLOWSET-EMBEDDED sample name (extra check rather than relying on pData)
  b$FileNo <- s  # file number according to flowframe index in flowset
  x <- pData(cs[s])  # extract flowset phenodata for sample n
  for(p in pDatalist){
    b[[p]] <- x[[p]]  # appends matching pData columns to flowframe dataframe
  }
  b <- as.data.table(b)  # converts dataframe to data.table
  data.list[[s]] <- b  # appends data.table to list, in order
}


```

## Check data
```{r}
## Data check
check <- do.list.summary(data.list)
check$name.table # Review column names and their subsequent values
check$ncol.check # Review number of columns (features, markers) in each sample
check$nrow.check # Review number of rows (cells) in each sample (should be same as number of cells in lymph gate for each sample)
data.list[[1]]  # view first and last six rows of data from first sample
```

## Data.table list to combined data.table
```{r}
## Data.list to Data.table ----
cell.dat <- Spectre::do.merge.files(dat = data.list)
cell.dat
```

# FlowSOM clustering
## Define cluster and analysis columns
```{r}
as.matrix(names(cell.dat))
cellular.cols <- names(cell.dat)[c(1, 4:7, 8:11)] #for MFI analyses, including MR1-tetramer
as.matrix(cellular.cols)
cluster.cols <- names(cell.dat)[c(1,4:7,9:13)]  # exclude live dead Zombie UV, CD3, MR1-tetramer
as.matrix(cluster.cols)

# exp.name <- 
sample.col <- "ID"
group.col <- "Condition"
# group.col2 <- "H_and_Y"
batch.col <- "EXPERIMENT_NAME"
```

## Run FlowSOM
```{r}
cell.dat <- run.flowsom(cell.dat, cluster.cols, meta.k = 5)  # meta.k specifies number of metaclusters
cell.dat
```

# Dimensionality reduction
## Downsampling
```{r}
## Set targets 
data.frame(table(cell.dat[[group.col]])) # Check number of cells/rows in each group
unique(cell.dat[[group.col]]) # check order groups appear in for subsequent assigning downsampling vector - in this case PD THEN control

sub.targets <- c(50000, 50000) # target subsample numbers from each group
sub.targets
# cell.sub <- do.subsample(cell.dat, divide.by= "ID", min.per=TRUE)


## Subsample groups by sub.targets
cell.sub <- do.subsample(cell.dat, sub.targets, group.col, seed=17)
cell.sub
```

## UMAP 
```{r}
cell.sub <- run.umap(cell.sub, cluster.cols)  # run DR (UMAP) on subsample using cluster.cols
cell.sub
```


# Cluster annotation
## Make plots to inform cluster annotation
```{r, warning=FALSE}
setwd("[working directory]")

# UMAP by cluster
make.colour.plot(cell.sub, "UMAP_X", "UMAP_Y", "FlowSOM_metacluster", col.type = 'factor',
                add.label=T, title = "Scaled parameters 300k cells")  # labels UMAP by cluster ID
# UMAP by marker density
make.multi.plot(cell.sub, "UMAP_X", "UMAP_Y", plot.by = cluster.cols, figure.title = "Marker Expression")  # plots individual UMAPs with overlaid marker expression (all markers, not just those used to cluster)

# cluster heatmap
exp <- do.aggregate(cell.dat, cluster.cols, by = "FlowSOM_metacluster")
make.pheatmap(exp, "FlowSOM_metacluster", cluster.cols, plot.title="Heatmap no annotation")
```

## Define annotations for MAIT
```{r}
annots <- list("CD4 SP MAIT"=1,
               "canonical MAIT"=2,
               "T-bet hi"=3, 
               "MR1-reactive T cell"=4,
               "CD45RO- MAIT"=5)

annots <- do.list.switch(annots)
names(annots) <- c("Values", "Population")
setorderv(annots, "Values")
annots

## Add to data.table and subsetted data.table ----
cell.dat <- do.add.cols(cell.dat, "FlowSOM_metacluster", annots, "Values")
cell.dat
cell.sub <- do.add.cols(cell.sub, "FlowSOM_metacluster", annots, "Values")
cell.sub
```



# PUBLICATION QUALITY ANNOTATED FIGS
## Batch integration check
```{r}
## Plot path
setwd("[working directory]")

## Batch integration check
make.multi.plot(cell.sub, "UMAP_X", "UMAP_Y", plot.by = 'EXPERIMENT_NAME', col.type = 'factor',
                plot.width = 20, figure.title="UMAP by Batch")  # overlays batches on single umap plot

make.multi.plot(cell.sub, "UMAP_X", "UMAP_Y", "Population", divide.by = 'EXPERIMENT_NAME',
                col.type = 'factor', plot.width = 20, figure.title="FlowSOM Metacluster by Batch")  # cluster-annotated umap plot for each batch to check presence of all clusters across batches

# UMAP with overlayed labelled metaclusters
make.colour.plot(cell.sub, "UMAP_X", "UMAP_Y", "Population", col.type = 'factor',
                 add.label = TRUE, nudge_x = -0.5, nudge_y = -0.5,
                 title = "", plot.height = 15, plot.width = 12,
                 dot.size = 2, blank.axis = F, legend.loc = 'none')  # labels UMAP by cluster ID
# no labels
make.colour.plot(cell.sub, "UMAP_X", "UMAP_Y", "Population", col.type = 'factor',
                 add.label = F,
                 plot.height = 15, plot.width = 12,
                 dot.size = 2, blank.axis = F, title="MAIT cells")  # labels UMAP by cluster ID

# UMAP plots by each marker used in clustering
make.multi.plot(cell.sub,"UMAP_X", "UMAP_Y", plot.by = cluster.cols, figure.title = "UMAP by Cluster Marker")

# UMAP plots by all markers used in clustering
make.multi.plot(cell.sub,"UMAP_X", "UMAP_Y", plot.by = cellular.cols, figure.title = "UMAP by Marker")

# UMAP plot divided by condition
make.multi.plot(cell.sub, "UMAP_X", "UMAP_Y", plot.by="Population", divide.by=group.col,
                col.type="factor", figure.title = "UMAP by Condition", dot.size=0.3)

# HEATMAP
exp <- do.aggregate(cell.dat, use.cols = cluster.cols, by = "Population")
make.pheatmap(dat = exp, sample.col = "Population", plot.cols = cluster.cols,
              dendrograms = "none", standard.colours = "BuPu",
              plot.title="")
```


# Summary data 
## Generate summary data
```{r, warning=FALSE}
as.matrix(names(cell.dat))  # check col names

## Generate summary data ----
# calculate percent of clusters in each sample (ID), as well as MFI of markers in each cluster
sum.dat <- create.sumtable(dat = cell.dat, # The dataset to be summarised
                           sample.col = 'ID', # The column that denotes the sample name/ID
                           pop.col = 'Population', # The column that denotes the population name/ID
                           use.cols = c("MR1-5-OP-RU PE", "T-bet BV605", "EOMES PE-Cy7"), # Columns (markers) whose expression we will measure on each population in each sample
                           annot.cols = c('Condition'), # Additional columns we would like to include for annotation purposes (e.g. group names, batch names, etc)
)


head(sum.dat)
```

# define inverse transform for MFIs
## inverse transform HLA-DR MFI
```{r}
trans <- gh_get_transformations(gs[[1]], inverse = TRUE)
trans
MR1trans <- trans[["Comp-[561] 582_15-A"]]  
Tbettrans <- trans[["Comp-[405] 610_20-A"]]
Eomestrans <- trans[["Comp-[561] 780_60-A"]]


MR1cols <- grep("MR1-5-OP-RU", colnames(sum.dat))
MR1cols
apply(sum.dat[,8:12], MARGIN = 2, MR1trans)  # check same as below
sum.dat[,8:12] <- lapply(sum.dat[,8:12], MR1trans)  
sum.dat[,8:12]

Tbetcols <- grep("BV605", colnames(sum.dat))
Tbetcols
apply(sum.dat[,13:17], MARGIN = 2, Tbettrans)  # check same as below
sum.dat[,13:17] <- lapply(sum.dat[,13:17], Tbettrans)  
sum.dat[,13:17] 

Eomescols <- grep("EOMES", colnames(sum.dat))
Eomescols
apply(sum.dat[,18:22], MARGIN = 2, Eomestrans)  # check same as below
sum.dat[,18:22] <- lapply(sum.dat[,18:22], Eomestrans)  
sum.dat[,18:22] 
```



## Convert NA population percentages into 0, edit sum.dat column names
```{r}
colnames(sum.dat)
percent.cols <- grep("Percent", colnames(sum.dat))  # defining plot columns as those that contain "percent"
for (i in percent.cols){
  index <- is.na(sum.dat[[i]])
  sum.dat[[i]][index] <- 0 # convert NA percentages, which based on function documentation indicate no population, to zero
}


colnames(sum.dat) <- gsub("Percent of sample -- ", "Percent.", colnames(sum.dat))
colnames(sum.dat) <- gsub("MFI of MR1-5-OP-RU PE -- ", "MR1.tet.MFI.", colnames(sum.dat))
colnames(sum.dat) <- gsub("MFI of T-bet BV605 -- ", "Tbet.MFI.", colnames(sum.dat))
colnames(sum.dat) <- gsub("MFI of EOMES PE-Cy7 -- ", "EOMES.MFI.", colnames(sum.dat))
colnames(sum.dat) <- gsub("CD45RO\\-", "CD45ROneg", colnames(sum.dat))

colnames(sum.dat) <- gsub(" ", ".", colnames(sum.dat))
colnames(sum.dat) <- gsub("\\-", ".", colnames(sum.dat))


colnames(sum.dat)

```


## Save Summary Data
```{r, warning=FALSE}
setwd("[working directory]")

PopSummaryData <- as.data.frame(sum.dat)
colnames(PopSummaryData)
write.csv(PopSummaryData, "MAIT_PopSummaryData.csv", row.names = F)  # entire dataframe including marker MFI on respective clusters
```


